What's hot and what's not:\\Windowed developer topic analysis

### 2009-06-23 lda-paper.tex
more ric edits

In this paper we explore how topics shift over time in the source
control system (SCS) of a project, using several open source database
systems as examples. We analyze commit log comments in each time
window and we extract the topics of that time window. We expect that
topics will change over time and the similarities and differences in
these topics will indicate developer focus and changes in developer
focus as it changes over time. We will show that most topics are
\hla{
locally relevant and thus a windowed topic analysis (local topic
analysis of windows of messages) can allow for both different kinds of
global and local analysis than global topic analysis would.
}

\hla{
Our contributions include:
\begin{itemize}
\item We demonstrate that windowed topic analysis most topics are local
  in scope, thus would be missed by a global topic analysis.
\item We present multiple visualizations of topics and trends over time to aid
  communication and analysis of topics and trends.
\item We provide an exploratory case study using these techniques on
  several open source database systems.
\end{itemize}
}

@@

We now define some terms that we will use in the rest of the paper: A
\emph{message} is a block of text written by developers. In this
paper, messages will be the CVS and BitKeeper commit log comments made
when the user commits changes to files in a repository. A \emph{word
  distribution} is the summary of a message by word count. Each word
distribution will be over the words in all messages. However, most
words will not appear in a message, a word distribution is effectively
a word count divided by the message size. A \emph{topic} is a word
distribution, a set of words that form a word distribution that is
unique and independent within the set of documents in our total
corpus. One could think of a topic as a distribution of the centroid
of a group of messages. In this paper we often summarize topics by the
top 10 most frequent words of their word distribution.  A \emph{trend}
is one or more similar topics that recur over time.  Trends are
particularly important, as they indicate long-lived and recurring
topics that may provide key insights into the development of the
project.

@@

In terms of clustering and finding topic distributions, Latent
Dirichlet Allocation (LDA)~\cite{944937} competes with Latent Semantic
Indexing
%(LSI)~\cite{1421013,1374321,10.1109/ICPC.2007.13,10.1109/ICPC.2006.17},
(LSI)~\cite{1374321,10.1109/ICPC.2007.13},
probabilistic Latent Semantic Indexing (pLSI)~\cite{944937} and
semantic clustering~\cite{1698774,1566153}. These tools are used for
document modelling, document clustering and collaborative
filtering. LDA attempts to encode documents as a mixture
model, a combination of topics.  LDA is used in software engineering
literature~\cite{lukins2008,10.1109/MSR.2007.20}%,NIPS2007637,1321709}
 to extract topics from documents such as methods, bug
reports, source code and files.

LDA, LSI and semantic clustering extract topic
clusters from the documents that are independent from one another. LDA
does not name these clusters but the clusters consist of words whose
relationship is often obvious to someone familiar with the corpus. We
could swap LDA for LSI or semantic clustering and likely produce similar
results. Our data is posed as documents with word distributions (word
counts per documents) and LDA extracts distinct topics (clusters of
words) from the documents.

@@

\hla{
Imagine that we are analyzing one month of development in a project
where the main focus was on bug fixing, adding a new GUI and
documentation updates relating to the system and the GUI. To use LDA
we'd get the word distributions of each commit messages, pass them to
LDA and tell LDA to find some topics, in this case we said 4. The 4
topics it discovered would probably be bug fixing, GUI, documentation
and other. These topics are essentially word distributions, for
example the bug fixing topic would include words like \emph{bug},
\emph{fix}, \emph{failure} and \emph{ticket}. The GUI topic would
include words like \emph{widget}, \emph{panel}, \emph{window}, and
\emph{menu}. The documentation topic would include words like
\emph{section}, \emph{figure}, and \emph{chapter}. The other topic
would include words found in topics generally independant of the other
topics. If we analyzed a commit of a bug fix it'd probably contain
words like \emph{ticket} and thus heavily associate with the bug fix
topic. A commit that was about user documentation regarding the GUI
would be associated with the documentation topic and the GUI topic
because it shared words from both topics. Where as a topic dealing
neither bugs, GUIs or documentation might be associated with the other
topic. This example demonstrates how LDA can associate 1 document (a
commit message) with more than one topic, how topics are related to
documents via their word distributions. LDA also would discover these
independant topics in an unsupervised manner, all the end user has to
do is name the topic based upon the most frequent words in that topic,
which is a word distribution.
}

@@

In our first exploratory pass we wanted to see if LDA could provide
interesting topics extracted from a real system. We took the
repository of MySQL 3.23, extracted the commits, grouped them by 30
day non-overlapping windows, and then applied LDA to each of these
windows. We asked LDA to make 20 topics per window and we then
\hla{
examined the top 10 most frequent words in that topic.  We chose one
month because it was smaller than the time between minor releases but
large enough for there to be many commits to analyze. We chose 20
topics because past experimentation showed that few topics might
aggregate multiple unique topics while anymore topics seemed to dilute
the results and create some indistinct topics. 10 most frequent words
were chosen as sometimes 1 common word dominated topics so we wanted
to have enough words to distingush topics.

We found there were common words across topic clusters, such as
\emph{diffs}, \emph{annotate} and \emph{history}; since these words
occur so often in the MySQL 3.23 messages. These words probably should
be viewed as stop words since occur across so many topics, even though
they are related to version control usage.  There were notable
transitional topics such as in the first window the word
\emph{BitKeeper} appears because MySQL adopted Bitkeeper for their
source control system yet in the following windows there were no
mentions of BitKeeper. \emph{RENAME} also only appeared once in the
first window and never again. Per each topic we tried to identify the
purpose of the topic; to our surprise we found that even with only a
little familiarity with the code base that naming the topic was
straightforward. To find the purpose of a commit we looked at the most
frequent words in the word distribution of the topic and tried to
summarize the topic, then we looked into the commits related to that
topic to investigate if we were correct; since the commit messages and
the word distribution share the same words the purpose extracted from
commit comments was usually accurate.
}

A sampling of the notable topic words is displayed in Table
\ref{tab:portability}, we chose topics that we felt confident we could
name.  To name each topic we selected a term that seemed to best
summarize that topic.  After extracting these topics, we attempted to
track the evolution of topics by visualizing the topics and joining
similar topics into trends.  Figure \ref{fig:lda} displays a manually
created sample plot of the extracted topics in Table
\ref{tab:portability}.

@@

The documents we are analyzing are the commit log comments i.e., the
comments that are added when revisions to the project are committed.
Per each commit log comment, we count the occurrence of each word in
the message, remove stop words, and then produce word distributions
for each message.  These distributions are then normalized by the size
of the message. After all messages are processed the distributions are
extended to include all words from all of the distributions.

@@

Given all messages, we group the messages into windows. We could use
overlapping windows, but in this paper we use non-overlapping windows
of a set time unit because it simplifies analysis.  Windowing by time
allows for many levels of granularity.  We used one month as the
length of our time windows. While we could use different window
lengths for this study we think that a month is a sizable unit of
development, which is large enough to show shifts in focus, but coarse
\hla{
enough not to show too much detail. Choosing a month as the window
provided us with enough documents to analyze. One could overlap the
windows, which would increase the likelihood of trends, but for this
study we lacked the space and were more concerned if topics ever
repeated, overlapping might skew that result.
}

>>>>

In this paper we explore how topics shift over time in the source
control system (SCS) of a project, using several open source database
systems as examples. We analyze commit log comments in each time
window and we extract the topics of that time window. We expect that
topics will change over time and the similarities and differences in
these topics will indicate developer focus and changes in developer
focus as it changes over time. We will show that most topics are
\hlb{
locally relevant, i.e., relevant to a single window of time.  
Our windowing approach supports both local (single window) analysis
and global (entire history) analysis.
}

\hla{
Our contributions include:
\begin{itemize}
\item We demonstrate that windowed topic analysis most topics are local
  in scope, which would be missed by a global topic analysis.
\item We present a number of visualizations of topics and their trends over time to aid
  communication and analysis of these topics and trends.
\item We provide an exploratory case study using these techniques on
  several open source database systems.
\end{itemize}
}

@@

We now define some terms that we will use in the rest of the paper: A
\emph{message} is a block of text written by developers. In this
paper, messages will be the CVS and BitKeeper commit log comments made
when the user commits changes to files in a repository. A \emph{word
  distribution} is the summary of a message by word count. Each word
distribution will be over the words in all messages. However, most
words will not appear in each message. A word distribution is effectively
a word count divided by the message size. A \emph{topic} is a word
distribution, i.e. a set of words that form a word distribution that is
unique and independent within the set of documents in our total
corpus. One could think of a topic as a distribution of the centroid
of a group of messages. In this paper we often summarize topics by the
top 10 most frequent words of their word distribution.  A \emph{trend}
is one or more similar topics that recur over time.  Trends are
particularly important, as they indicate long-lived and recurring
topics that may provide key insights into the development of the
project.

@@

In terms of clustering and finding topic distributions, Latent
Dirichlet Allocation (LDA)~\cite{944937} competes with Latent Semantic
Indexing
%(LSI)~\cite{1421013,1374321,10.1109/ICPC.2007.13,10.1109/ICPC.2006.17},
(LSI)~\cite{1374321,10.1109/ICPC.2007.13},
probabilistic Latent Semantic Indexing (pLSI)~\cite{944937} and
semantic clustering~\cite{1698774,1566153}. These tools are used for
document modelling, document clustering and collaborative
filtering. LDA attempts to encode documents as a mixture
model, a combination of topics.  LDA is used in software engineering
literature~\cite{lukins2008,10.1109/MSR.2007.20}%,NIPS2007637,1321709}.
LDA extracts topics from documents such as methods, bug
reports, source code and files.

LDA, LSI and semantic clustering extract topic
clusters from the documents that are independent from one another. LDA
does not name these clusters but the clusters consist of words whose
relationship is often obvious to someone familiar with the corpus. For our purposes, we
could swap LDA for LSI or semantic clustering and likely produce similar
results. Our data is posed as documents with word distributions (word
counts per documents) and LDA extracts distinct topics (clusters of
words) from the documents.

@@

\hla{
Imagine that we are analyzing one month of development in a project
where the main focus was on bug fixing, adding a new GUI and
documentation updates relating to the system and the GUI. To use LDA
we would get the word distributions of each commit messages, pass them to
LDA and tell LDA to find some topics, in this case we
\hlb{
 in this case we chose to concentrate on four topics.
} The four
topics it discovered would ideally be bug fixing, GUI, documentation
and other. These topics are essentially word distributions, for
example the bug fixing topic would include words like \emph{bug},
\emph{fix}, \emph{failure} and \emph{ticket}. The GUI topic would
include words like \emph{widget}, \emph{panel}, \emph{window}, and
\emph{menu}. The documentation topic would include words like
\emph{section}, \emph{figure}, and \emph{chapter}. The other topic
would include words found in topics generally independant of the other
topics. If we analyzed a commit of a bug fix it would probably contain
words like \emph{ticket} and thus heavily associate with the bug fix
topic. A commit that was about user documentation regarding the GUI
would be associated with the documentation topic and the GUI topic
because it shared words from both topics. Where as a topic dealing
neither bugs, GUIs or documentation might be associated with the other
topic. This example demonstrates how LDA can associate one document (a
commit message) with more than one topic, and how topics are related to
documents via their word distributions. LDA also would discover these
independant topics in an unsupervised manner. All the end user has to
do is name the topic based upon the most frequent words in that topic,
which is a word distribution.
}

@@

In our first exploratory pass we wanted to see if LDA could provide
interesting topics extracted from a real system. We took the
repository of MySQL 3.23, extracted the commits, grouped them by 30
day non-overlapping windows, and then applied LDA to each of these
windows. We asked LDA to make 20 topics per window and we then
\hla{
examined the top 10 most frequent words in that topic.  We chose one
month because it was smaller than the time between minor releases but
large enough for there to be many commits to analyze. We chose 20
topics because past experimentation showed that fewer topics might
aggregate multiple unique topics while any more topics seemed to dilute
the results and create indistinct topics. 
\hlb{
  We chose the top ten most frequent words because sometimes a couple
  of common words dominated topics so we wanted to have enough words
  to distinguish topics.
}

We found there were common words across topic clusters, such as
\emph{diffs}, \emph{annotate} and \emph{history}; since these words
occur so often in the MySQL 3.23 messages. These words probably should
be viewed as stop words since occur across so many topics, even though
they are related to version control usage.  There were notable
transitional topics such as in the first window the word
\emph{BitKeeper} appears because MySQL adopted Bitkeeper for their
source control system yet in the following windows there were no
mentions of BitKeeper. \emph{RENAME} also only appeared once in the
first window and never again. For each topic we tried to name the
purpose of the topic; to our surprise we found that even with only a
little familiarity with the code base that naming the topic was
straightforward. To find the purpose of a commit we looked at the most
frequent words in the word distribution of the topic and tried to
summarize the topic, then we looked into the commits related to that
topic to investigate if we were correct; since the commit messages and
the word distribution share the same words the purpose extracted from
commit comments was usually accurate.
}

A sampling of the notable topic words is displayed in Table
\ref{tab:portability}, we chose topics that we felt confident we could
name.  To name each topic we selected a term that seemed to best
summarize that topic.  After extracting these topics, we attempted to
track the evolution of topics by visualizing the topics and joining
similar topics into trends.  Figure \ref{fig:lda} displays a manually
created plot of the extracted topics in Table
\ref{tab:portability}.

@@

The documents we are analyzing are the commit log comments, i.e. the
comments that are added when revisions to the project are committed.
For each commit log comment, we count the occurrence of each word in
the message, remove stop words, and then produce word distributions
for each message.  These distributions are then normalized by the size
of the message, each count was divided by the total number of
tokens. After all messages are processed the distributions are
extended to include all words from all of the distributions.

@@

Given all messages, we group the messages into windows. We could use
overlapping windows, but in this paper we use non-overlapping windows
of a set time unit because it simplifies analysis.  
One could overlap the
windows, which would increase the likelihood of trends, but for this
study we lacked the space and were more concerned if topics ever
repeated, overlapping might skew that result.
Windowing by time
allows for many levels of granularity.  We used one month as the
length of our time windows. While we could use different window
lengths for this study we think that a month is a sizable unit of
development, which is large enough to show shifts in focus, but coarse
\hla{
enough not to show too much detail. Choosing a month as the window
provided us with enough documents to analyze. 
}
